---
# title: "Manuscript Analysis"
format:
  html:
    toc: true
    toc-expand: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    self-contained: true
    code-fold: false
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
  message: false
  echo: true
---

# Project Setup

```{r}
#| label: setup

source('https://inkaverse.com/setup.r')
```

# Data import

```{r}
gs <- "https://docs.google.com/spreadsheets/d/1cNoexwoni9u8OxOOeWBAZIlYe6dcXQ4yPVuuHt_pOsU/edit?gid=0#gid=0" %>% 
  as_sheets_id()

fb <- gs %>% 
  range_read(ss = ., sheet = "fb") %>% 
  mutate(
    variedades = factor(variedades),
    rep = factor(rep),
    planta = factor(planta)
  )

str(fb)

```

```{r}
library(tidyverse)

ggplot(fb, aes(x = variedades, y = severidad)) +
  geom_boxplot() +
  labs(
    x = "Variedad",
    y = "Severidad"
  ) +
  theme_classic()
```

#ANOVA
```{r}
av <- aov(severidad ~ variedades, fb)
anova(av)
plot(av)
mc <-  emmeans(av, ~ variedades) %>%
  cld(Letters = letters)

mc
```

#GRAFICOS
```{r}
ggplot(mc, aes(x = variedades, y = emmean)) +
  geom_col(aes(fill = variedades)) +
  geom_text(
    aes(label = .group),
    vjust = -0.5,
    size = 5
  ) +
  scale_y_continuous(
    limits = c(0, 30),
    breaks = seq(0, 30, 5),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Variedades",
    y = "Severidad",
    fill = "Variedades"
  ) +
  theme_classic()
```

## RO 2

```{r}

```

# Multivariate

```{r}
#| eval: false

library(FactoMineR)
library(factoextra)

dtx <- fb %>% 
  dplyr::select(!c("var1", "var2")) %>% 
  as.data.frame()
# str(dtx)
  
mv <- dtx %>% 
  PCA(scale.unit = T
      , graph = F
      , quali.sup = c(1, 2))

var <- mv %>% 
  plot.PCA(choix = c("var"), cex=0.7)

cos <- mv %>% 
  fviz_pca_var(col.var = "cos2"
               , gradient.cols = inti:::paleta()
               , repel = TRUE
               , labelsize = 2)

ind <- mv %>% 
  plot.PCA(choix = c("ind")
           , cex = 0.7
           , label = "ind"
           , invisible = "quali"
           , col.hab = inti:::paleta()
           , habillage = "none"
           )

pcainfo <- factoextra::get_pca(mv); pcainfo

pcainfo$cor %>% kable(caption = "Variable correlations")
pcainfo$contrib %>% kable(caption = "Variable Contribution")
pcainfo$cos2 %>% kable(caption = "Variable Cos2")

cls <- mv %>% HCPC(nb.clust = -1, graph = FALSE)

map <- \() {plot.HCPC(cls, choice = "map", draw.tree = FALSE)}

cls$data.clust # Cluster membership of each individual
cls$desc.var$quanti # Quantitative variables characterizing each cluster
cls$desc.var$category # Categorical variables significantly by cluster 
table(cls$data.clust$clust) # Number of individuals assigned to each cluster

plot <- list(var, cos, ind, map) %>% 
  plot_grid(plotlist = ., ncol = 2, labels = "auto") %>% 
  ggsave(plot = ., "manuscript/multivariate.jpg", units = "cm"
                , width = 30, height = 30)

plot %>% include_graphics()
```

# References

Broman, K. W., & Woo, K. H. (2017). Data organization in spreadsheets. The American Statistician, 72(1), 2–10. https://doi.org/10.1080/00031305.2017.1375989

Zuur, A. F., Ieno, E. N., & Elphick, C. S. (2009). A protocol for data exploration to avoid common statistical problems. Methods in Ecology and Evolution, 1(1), 3–14. https://doi.org/10.1111/j.2041-210x.2009.00001.x

Francoishusson. (2017, July 13). PCA course using FactoMineR | R-bloggers. R-bloggers. https://www.r-bloggers.com/2017/07/pca-course-using-factominer/

Kozak, M., & Piepho, H. (2017). What’s normal anyway? Residual plots are more telling than significance tests when checking ANOVA assumptions. Journal of Agronomy and Crop Science, 204(1), 86–98. https://doi.org/10.1111/jac.12220

Tanaka, E., & Hui, F. K. C. (2019). Symbolic formulae for linear mixed models. In Communications in computer and information science (pp. 3–21). https://doi.org/10.1007/978-981-15-1960-4_1

Schielzeth, H., Dingemanse, N. J., Nakagawa, S., Westneat, D. F., Allegue, H., Teplitsky, C., Réale, D., Dochtermann, N. A., Garamszegi, L. Z., & Araya‐Ajoy, Y. G. (2020). Robustness of linear mixed‐effects models to violations of distributional assumptions. Methods in Ecology and Evolution, 11(9), 1141–1152. https://doi.org/10.1111/2041-210x.13434

Jambor, H. K. (2025). A checklist for designing and improving the visualization of scientific data. Nature Cell Biology, 27(6), 879–883. https://doi.org/10.1038/s41556-025-01684-z
